Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/appendix.tex
more comments
work on appendix

There are two relevant notions of \emph{morphism} of CBPV models:
\emph{strict} and \emph{lax}.
% morphisms of CBPV models
Given call-by-push-value models
$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
that strictly presere the constructors:
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) = 1_2$
\end{enumerate}
As well as strictly preserving the corresponding universal morphisms
and coherence isomorphisms.

@@

\section{Details of the Construction of an Extensional Model}

In Section \ref{sec:extensional-model-construction}, we outline the construction
of an extensional model of gradual typing starting from a step-1 intensional model.
In this section, we provide the details for each of the constructions mentioned there.

\subsection{Constructing a Model with Perturbations}

The goal of this section is to prove the following lemma:

\begin{lemma}\label{lem:step-1-model-to-step-2-model}
Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.

Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
\end{lemma}

We begin with a definition and some lemmas that will be useful in the construction
of the model.

\begin{definition}
  Let $c : A \rel A'$ be a value relation of $\mathcal M$.
  Let $P_A$ be a monoid of perturbations on $A$ and $P_{A'}$ a monoid of perturbations on $A'$.
  A \emph{push-pull structure} $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$ consists of:
  \begin{itemize}
    \item A function $\push : P_A \to P_{A'}$ 
    such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
    \item A function $\pull : P_{A'} \to P_A$
    such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
  \end{itemize}

  We define a push-pull structure $\pie_d$ for $d : B \rel B'$ with respect to perturbation monoids
  $P_B$ and $P_{B'}$ in an analogous manner.

  % \eric{do we need that $\delta^*$ is preserved by push and pull?}
  % We stipulate the additional requirement that if $\delta_A^*$ is in $P_B$ 
  % (where $\delta_A^*$ is the distinguished perturbation required to be in every $\ef(FA, FA)$),
  % then $\push(\delta_A^*) = \delta_{A'}^*$ and likewise $\pull(\delta_{A'}^*) = \delta_A^*$.
\end{definition}

%%%%%%%%%%%%%%%
% Composition %
%%%%%%%%%%%%%%%
\begin{lemma}\label{lem:push-pull-comp}
  Let $c : A \rel A'$ and $c' : A' \rel A''$ be value relations,
  and let $P_A, P_{A'}$ and $P_{A''}$ be monoids of perturbations.
  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$, and
  and a push-pull structure $\piv_{c'}$ for $c'$ with respect to $P_{A'}$ and $P_{A''}$,
  we can define a push-pull structure $\piv_{c \comp c'}$ for $c \comp c'$.

  Likewise, we can define a push-pull structure for the composition of computation relations.
\end{lemma}
\begin{proof}
  We define $\piv_{c \comp c'}$ as the following push-pull structure:

  \begin{itemize}
    \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
    \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
  \end{itemize}

  We observe that the required squares exist for both push and pull.
  In particular, for push we have that $\delta^l \ltdyn_c^c \push_c(\delta^l)$
  using the push property for $c$, and then using the push property for $c'$ we have
  $\push_c(\delta^l) \ltdyn_{c'}^{c'} \push_{c'}(\push_c(\delta^l))$.
  We can then compose these squares horizontally to obtain the desired square.
  The pull property follows similarly.

@@

%%%%%%%%%%%
% U and F %
%%%%%%%%%%%
\begin{lemma}\label{lem:push-pull-U-F}
  Let $c : A \rel A'$ and let $P_A$ and $P_{A'}$ be monoids of perturbations.
  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$,
  we can define a push-pull structure $\pie_{Fc}$ for $F(c)$ with respect to
  $\mathbb{N} \times P_{A}$ and $\mathbb{N} \times P_{A'}$.

  Likewise, from a push-pull structure for $d$ with respect to $P_B$ and $P_{B'}$,
  we can define a push-pull structure for $Ud$ with respect to $\mathbb{N} \times P_B$
  and $\mathbb{N} \times P_{B'}$.
\end{lemma}
\begin{proof}

  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$ by

  \[ \push_{Fc}(n, a) = (n, \push_c(a)). \]

  Then we need to build the following square:

  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGQSJdLFsxLDAsIkZBJyJdLFswLDEsIkZBIl0sWzEsMSwiRkEnIl0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCIoXFxkZWx0YV97RkF9XiopXm4gXFxjaXJjIFxccHRiX0EoYSkiLDJdLFsxLDMsIihcXGRlbHRhX3tGQSd9XiopXm4gXFxjaXJjIFxccHRiX3tBJ30oXFxwdXNoX2MoYSkpIl1d
  \[\begin{tikzcd}[ampersand replacement=\&]
    FA \& {FA'} \\
    FA \& {FA'}
    \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \arrow["Fc", "\shortmid"{marking}, no head, from=1-1, to=1-2]
    \arrow["{(\delta_{FA}^*)^n \circ \ptb_A(a)}"', from=1-1, to=2-1]
    \arrow["{(\delta_{FA'}^*)^n \circ \ptb_{A'}(\push_c(a))}", from=1-2, to=2-2]
  \end{tikzcd}\]

  This is obtained by composing the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
  with the square corresponding to the push property for $B$

  % Given $n \in \mathbb{N}$ we define the push to be $n$.
  % Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$,
  % which exists by the definition of a step-1 model.
  % On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
  % and then the required exists by the push property for $c$.
  % Defining $\pull_{Fc}$ is symmetric.

%%%%%%%%%%%%
% Products %
%%%%%%%%%%%%
\begin{lemma}\label{lem:push-pull-times}
  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, and let $P_{A_1}$,
  $P_{A_2}$, $P_{A_1'}$ and $P_{A_2'}$ be monoids of perturbations.
  Given push-pull structures 
  $\piv_{c_1}$ with respect to $P_{A_1}$ and $P_{A_1'}$, and 
  $\piv_{c_2}$ with respect to $P_{A_2}$ and $P_{A_2'}$,
  we can define a push-pull structure $\piv_{c_1 \times c_2}$
  with respect to $P_{A_1} \times P_{A_2}$ and $P_{A_1'} \times P_{A_2'}$.

  % Likewise, given push-pull structures 
  % $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
  % $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
  % we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
  % with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
\end{lemma}
\begin{proof}
We define the push function for $\piv_{c_1 \times c_2}$ by 
$\push(a_1, a_2) = (\push_{c_1}(a_1), \push_{c_2}(a_2))$ and likewise for $\pull$.
The push property holds because it holds for each component.
The pull property is established similarly.
\end{proof}

%%%%%%%%%
% Arrow %
%%%%%%%%%
\begin{lemma}\label{lem:push-pull-arrow}
  Let $c : A \rel A'$ and $d : B \rel B'$.
  Given push-pull structures $\piv_c$ for $c$ and $\pie_d$ for $d$,
  we can define a push-pull structure $\pie_{c \arr d}$ for $c \arr d$.

  % Given push-pull structures $\pie_{Fc}$ for $Fc$ and $\piv_{Ud}$ for $Ud$,
  % we can define a push-pull structure $\piv_{U(c \arr d)}$ for $U(c \arr d)$
  % with respect to the monoids $(P_{FA})^{op} \oplus P_B$
\end{lemma}
\begin{proof}
  Follows from the functorial action of $\arr$ on perturbations and squares.
  % and the latter uses the functorial action of $\tok$ on perturbations and squares.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%
% Model Construction %
%%%%%%%%%%%%%%%%%%%%%%
We now proceed with the construction of the model:

    Define a step-2 model $\mathcal M'$ as follows:
    \begin{itemize}
      \item Value objects are pairs consisting of:
      \begin{itemize}
        \item A value object $A$ in $\vf$.
        \item A commutative monoid $P_A$ of perturbations
        and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.

      \item Computation objects are pairs consisting of:
      \begin{itemize}
        \item A computation object $B$ in $\ef$.
        \item A commutative monoid $P_B$ of perturbations and a monoid homomorphism
        $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
      \end{itemize}

      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively.
      %, i.e.,
      % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
      %
      % and likewise for computations.

      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
      \begin{itemize}
        \item A value relation $c \in \vsq$.
        \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
        %\item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
      \end{itemize}

      The objects of $\esq'$ are defined analogously, i.e.,
      \begin{itemize}
        \item A computation relation $d \in \esq$.
        \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
        %\item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
      \end{itemize}

      \item The morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.

      \item We define composition of relations $(c, \piv_c)$ and $(c', \piv_{c'})$
      as $(c \comp c', \piv_{c \comp c'})$ where $\piv_{c \comp c'}$ is as in Lemma
      \ref{lem:push-pull-comp}, and likewise for computation relations.

      % Functors \times, +, F, U, arrow
      Now we define the actions of the functors:
      \begin{itemize}

      \item We define $\times$ on objects by

      \[ (A_1, P_{A_1}, \ptb_{A_1}) \times (A_2, P_{A_2}, \ptb_{A_2}) =
        (A_1 \times A_2, P_{A_1} \times P_{A_2}, \ptb_{A_1 \times A_2}) \]

      where $\ptb_{A_1 \times A_2}(p_1, p_2) = \ptb_{A_1}(p_1) \times \ptb_{A_2}(p_2)$.

      Using Lemma \ref{lem:push-pull-times}, we define $\times$ on relations by

      \[ (c_1, \piv_{c_1}), (c_2 \piv_{c_2}) = 
        (c_1 \times c_2, \piv_{c_1 \times c_2}). \]

      \item We define $F$ on objects by

      \[ F(A, P_A, \ptb_A) = (FA, \mathbb{N} \times P_A, \ptb_{FA}), \]

      where we define $\ptb_{FA}(n, a) = (\delta_{FA}^*)^n \circ F(\ptb_A(a))$.

      Using Lemma \ref{lem:push-pull-U-F}, we define $F$ on relations by

      \[ F(c, \piv_c) = (Fc, \pie_{Fc}). \]

      \item We define $U$ on objects by

      \[ U(B, P_B, \ptb_B) = (UB, \mathbb{N} \times P_B, \ptb_{UB}), \]

      where we define $\ptb_{UB}(n, b) = (\delta_{UB}^*)^n \circ U(\ptb_B(b))$.

      Using Lemma \ref{lem:push-pull-U-F}, we define $U$ on relations by

      \[ U(d, \pie_d) = (Ud, \piv_{Ud}).  \]

      We define $\arr$ on objects by

      \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
        (A \arr B, P_A^{op} \times P_B, \ptb_{A \arr B}), \]

      where we define $\ptb_{A \arr B}(a, b) = \ptb_A(a) \arr \ptb_B(b)$.

      Using Lemma \ref{lem:push-pull-arrow}, we define $\arr$ on relations by

      \[ (c, \piv_c) \arr (d, \pie_d) =
        (c \arr d, \pie_{c \arr d}). \]

    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
    to a perturbation in $P_{U(A \to B)}$, as follows:
    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
    We need to show that
%
    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
%
    That is, we need to show that
%
    \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \ptb_B(b)) = 
        \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
%
    This follows from the fact that $\delta_{UB}^*$ commutes with computation morphisms.
%

    Given a perturbation $q = (n, a) \in P_{FA}$ we define
    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
    We need to show that
  %
    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
  %
    By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes with all computation morphisms.

    We can similarly construct a Kleisli product operation on perturbations.
    Given a perturbation $(n, a_1, a_2) \in P_{F(A_1 \times A_2)}$ we define
    $(n, a_1, a_2) \timesk \id = (n, a_1, \id_{P_{A_2}})$,
    and likewise we define $\id \timesk (n, a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.

@@

>>>>

\max{there is not definition of a CBPV model...}
There are two relevant notions of \emph{morphism} of CBPV models:
\emph{strict} and \emph{lax}.
% morphisms of CBPV models
Given call-by-push-value models
$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
that strictly presere the constructors:
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) = 1_2$
\end{enumerate}
As well as strictly preserving the corresponding universal morphisms
and coherence isomorphisms.

@@

\section{Omitted Proofs Section \ref{sec:towards-relational-model}}

\begin{theorem}
  Let $R$ be a binary relation on the free error domain $U(\li A)$. If
  $R$ satisfies the following properties
  \begin{enumerate}
  \item Transitivity
  \item $\theta$-congruence: If $\later_t (\tilde{x}_t \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R} \theta(\tilde{y})$.
  \item Right step-insensitivity: If $x \binrel{R} y$ then $x \binrel{R} \delta y$.
  \end{enumerate}
  Then for any $l : U(\li A)$, we have $l \binrel{R} \Omega$. If $R$ is left
  step-insensitive instead then $\Omega \binrel{R} x$.
\end{theorem}
\begin{proof}
  By \lob-induction: we assume that $\laterhs (l \binrel{R} \Omega)$, and we
  show $(l \binrel{R} \Omega)$. We have $l \binrel{R} \delta l$ by right
  step-insensitivity.
  %
  By transitivity, it will therefore suffice to show that $\delta l \binrel{R}
  \Omega$.
  %
  Then since $\Omega = \delta \Omega$, it suffices to show $\delta l \binrel{R}
  \delta\Omega$.
  %
  By $\theta$-congruence, it suffices to show that
  $\later_t [(\nxt\, l)_t \binrel{R} (\nxt\, \Omega)_t]$, i.e., $\later (l \binrel{R} \Omega)$,
  which is our induction hypothesis.

  The proof when $R$ is left step-insensitive is symmetric.
\end{proof}

\section{Details of the Relational Model Construction}

In Section \ref{sec:concrete-relational-model} we described our model of gradual
typing. We omitted several technical proofs involving the well-definedness of
composition of relations and the actions of functors. We provide those proofs in
this section.

% We note that our approach in this section will be differ from the body of the
% paper in two ways:

% First, we will work \emph{abstractly} with a notion of model of gradual typing
% defined in the setting of an arbitrary double category.
% %
% Second, we build up the abstract notion of model in several phases, and layer
% the constructions in a modular way so that we can isolate the essential parts of
% the proofs and abstract away unnecessary details about previous parts of the
% construction.

% The constructions we carry out in this section specialize to the model we
% describe in the body of the paper. We find, however, that organizing things in
% this manner makes the arguments easier to follow.

\subsection{Free Composition of Error Domain Relations}
\begin{definition}\label{def:free-comp-ed-rel}
  Let $d : B_1 \rel B_2$ and $d' : B_2 \rel B_3$ be error domain relations. We
  define the composition $dd'$ inductively by the following rules:
  \begin{mathpar}
      \inferrule*[right = Comp]
      {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
      {b_1 \mathbin{d \relcomp d'} b_3}

      \inferrule*[right = DnClosed]
      {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
      {b_1' \mathbin{d \relcomp d'} b_3}

      \inferrule*[right = UpClosed]
      {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
      {b_1 \mathbin{d \relcomp d'} b_3'}

      \inferrule*[right = PresErr]
      { }
      {\mho_{B_1} \mathbin{d \relcomp d'} b_3}

      \inferrule*[right = PresTheta]
      {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
      {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
  \end{mathpar}
\end{definition}

\subsection{The Dynamic Type}
The ordering $\le_D$ on the predomain $D$ is defined by
\begin{align*}
  \inat(n) \le \inat(n') 
      &\iff n = n' \\
  \itimes (d_1, d_2) \le \itimes (d_1', d_2')
      &\iff d_1 \le_D d_2 \text{ and } d_1' \le_D d_2'\\
  \iarr(\tilde{f}) \le \iarr(\tilde{f'}) 
      &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t),
\end{align*}

and the bisimilarity relation is analogous.

\subsection{Lemmas about Perturbations}

The goal of this section is to prove the following lemmas:

%%%%%%%%%%%%%%%
% Composition %
%%%%%%%%%%%%%%%
\begin{lemma}\label{lem:push-pull-comp}
  Let $(A_1, M_{A_1}, i_{A_1})$, $(A_2, M_{A_2}, i_{A_2})$, and $(A_3, M_{A_3},
  i_{A_3})$ be value objects, and let $c : A_1 \rel A_2$ and $c' : A_2 \rel A_3$
  be predomain relations.

  Given a push-pull structure $\Pi_c$ for $c$ and $\Pi_{c'}$ for $c'$, we can
  define a push-pull structure $\Pi_{c \comp c'}$ for $c \comp c'$.

  Likewise, we can define a push-pull structure for the composition of error
  domain relations.
\end{lemma}
\begin{proof}
  We define $\piv_{c \comp c'}$ as follows: We first define
  %
  $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$ and 
  $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$.
  %
  We then observe that the required squares exist for both push and pull. In
  particular, for push we have that $i_{A_1}(m_{A_1}) \ltdyn_c^c
  i_{A_2}(\push_c(m_{A_1}))$ using the push property for $c$, and then using the
  push property for $c'$ we have $i_{A_2}(\push_c(m_{A_1})) \ltdyn_{c'}^{c'}
  i_{A_3}(\push_{c'}(\push_c(m_{A_1})))$. We can then compose these squares
  horizontally to obtain the desired square. The pull property follows
  similarly.

@@

%%%%%%%%%%%
% U and F %
%%%%%%%%%%%
\begin{lemma}\label{lem:push-pull-U-F}

  Let $A$ and $A'$ be value objects and let $c : A \rel A'$ be a relation on the
  underlying predomains. Given a push-pull structure $\Pi_c$ for $c$ we can
  define a push-pull structure $\Pi_{\li c}$ for $\li c$ with respect to $\li A$
  and $\li A'$.

  Likewise, let $B$ and $B'$ be computation objects and let $d : B \rel B'$ be
  an error domain relation. Given a push-pull structure for $d$ with respect to
  $B$ and $B'$, we can define a push-pull structure for $Ud$ with respect to
  $UB$ and $UB'$.
\end{lemma}
\begin{proof}

  We define $\push_{\li c} : \mathbb{N} \oplus M_{A} \to \mathbb{N} \oplus
  M_{A'}$ by the universal property of the coproduct of monoids. In particular,
  it suffices to define a homomorphism of monoids $\mathbb{N} \to \mathbb{N}
  \oplus M_{A'}$ and $M_{A} \to \mathbb{N} \oplus M_{A'}$.

  The former is simply $\inl$ and the latter is $\inr \circ \push_c$.

  Then to establish the push property, it suffices by the universal property of
  the coproduct, and the fact that $\mathbb{N}$ is the free monoid on one
  generator, to show that the following two squares exist:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEEiXSxbMSwwLCJcXGxpIEEnIl0sWzAsMSwiXFxsaSBBIl0sWzEsMSwiXFxsaSBBJyJdLFswLDEsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxsaSBjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJcXGRlbHRhXipfQSIsMl0sWzEsMywiXFxkZWx0YV4qX3tBJ30iXV0=
\[\begin{tikzcd}[ampersand replacement=\&]
	{\li A} \& {\li A'} \\
	{\li A} \& {\li A'}
	\arrow["{\li c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\delta^*_A}"', from=1-1, to=2-1]
	\arrow["{\delta^*_{A'}}", from=1-2, to=2-2]
	\arrow["{\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEEiXSxbMCwxLCJcXGxpIEEiXSxbMSwwLCJcXGxpIEEnIl0sWzEsMSwiXFxsaSBBJyJdLFswLDIsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiXFxsaSBjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGxpKGlfQShtX0EpKSIsMl0sWzIsMywiXFxsaSAoaV97QSd9KFxccHVzaF9jKG1fQSkpKSJdXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	{\li A} \& {\li A'} \\
	{\li A} \& {\li A'}
	\arrow["{\li c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\li(i_A(m_A))}"', from=1-1, to=2-1]
	\arrow["{\li (i_{A'}(\push_c(m_A)))}", from=1-2, to=2-2]
	\arrow["{\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

where $\delta* = (\delta \circ \eta)^\dagger$

The first square exists by the fact that the monadic extension operation
$-^\dagger$ is monotone in its argument, and the fact that there is a
square $\delta_A \circ \eta_A \ltsq{}{} \delta_{A'} \circ \eta_{A'}$.

The second square exists by the push property for $c$ and the fact that $\li$
acts on squares.

The proof for the pull property is dual, and the construction of a push-pull
structure for $Ud$ is analogous.
\end{proof}

%%%%%%%%%%%%
% Products %
%%%%%%%%%%%%
\begin{lemma}\label{lem:push-pull-times} Let $A_1$, $A_1'$, $A_2$, and $A_2'$ be
  value objects. Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$ be
  relations on the underlying predomains. Given push-pull structures $\Pi_{c_1}$
  for $c_1$ and $\Pi_{c_2}$ for $c_2$ we can define a push-pull structure
  $\Pi_{c_1 \times c_2}$ for $c_1 \times c_2$.
\end{lemma}
\begin{proof}
Recall that the monoid of syntactic perturbations for $A_1 \times A_2$ is
$M_{A_1} \oplus M_{A_2}$ and the monoid for $A_1' \times A_2'$ is $M_{A_1'}
\oplus M_{A_2'}$. Thus to define the push homomorphism for $c_1 \times c_2$, it
suffices by the universal property of the coproduct of monoids to define a
homomorphism $M_{A_1} \to M_{A_1'} \oplus M_{A_2'}$ and $M_{A_2} \to M_{A_1'}
\oplus M_{A_2'}$. The former is $\inl \circ \push_{c_1}$, and the latter is
$\inr \circ \push_{c_2}$. The pull homomorphism is defined similarly.

For the push property, we need to show that there is a square 
$i_{A_1 \times A_2}(m_1) \ltsq{}{} i_{A_1' \times A_2'}(\push_{c_1 \times c_2}(m_1))$.
%
By the universal property of the coproduct, it suffices to consider two cases: a
perturbation $m_1 \in M_{A_1}$ and a perturbation $m_2 \in M_{A_2}$. In the
first case, using the definition of the interpretation homomorphism for the
product, the square becomes $(i_{A_1}(m_1) \times \id_{A_2}) \ltsq{}{}
(i_{A_1'}(\push_{c_1}(m_1)) \times \id_{A_2'})$.
%
Then by the action of $\times$ on squares, it suffices to show that there are
squares $i_{A_1}(m_1) \ltsq{}{} i_{A_1'}(\push_{c_1}(m_1))$ and $\id_{A_2}
\ltsq{}{} \id_{A_2'}$. The former is the push property for $c_1$, and the latter
is an identity square.

The second case, i.e., $m_2 \in M_{A_2}$, is symmetric, and the proof of the
pull property is dual.
\end{proof}

%%%%%%%%%
% Arrow %
%%%%%%%%%
\begin{lemma}\label{lem:push-pull-arrow} Let $A$ and $A'$ be value objects and
  $B$ and $B'$ be computation objects. Let $c : A \rel A'$ be a relation on the
  underlying predomains and $d : B \rel B'$ a relation on the underlying error
  domains. Given push-pull structures $\Pi_c$ for $c$ and $\Pi_d$ for $d$, we
  can define a push-pull structure $\Pi_{c \arr d}$ for $c \arr d$.
\end{lemma}
\begin{proof}
  Similar to the product case above.
\end{proof}

We next define the actions of the Kleisli arrow and product functors on
syntactic perturbations.

% ``Kleisli arrow'' action on syntactic perturbations that takes a
% perturbation in $M_{UB}$ to a perturbation in $M_{U(A \to B)}$, as follows:

\begin{definition}\label{def:kleisli-arrow-perturbations} Let $A$ be a value
  object and $B$ be a computation object. We define a homomorphism of monoids
  $id \tok - \colon M_{UB} \to M_{U(A \arr B)}$ via the universal property of the
  coproduct of monoids by sending $\mathbb{N}$ to the first injection in $M_{U(A
  \arr B)} = \mathbb{N} \oplus M_A^{op} \oplus M_B$, and sending $M_B$ to the
  third injection.

  Likewise, we define $- \tok \id \colon M_{\li A}^{op} \to M_{U(A \arr B)}$ as
  follows. First note that $M_{\li A}^{op} = \mathbb{N} \oplus M_A^{op}$. Then
  we construct the homomorphism via the universal property, sending $\mathbb{N}$
  to the first injection $M_A^{op}$ to the second injection.
\end{definition}

We can similarly construct a Kleisli product operation on perturbations:

\begin{definition}\label{def:kleisli-product-perturbations}
  Let $A_1$ and $A_2$ be value objects. We define a homomorphism of monoids 
  $- \timesk \id \colon M_{\li A_1} \to M_{\li (A_1 \times A_2)}$ using the universal property,
  sending $\mathbb{N}$ to the first injection and $M_{A_1}$ to the second injection.

  Likewise, we define $\id \timesk - \colon M_{\li A_2} \to M_{\li (A_1 \times A_2)}$
  by sending $\mathbb{N}$ to the first injection and $M_{A_2}$ to the third injection.

  %  Given a perturbation $(n, a_1, a_2) \in $ we define $(n, a_1, a_2)
  % \timesk \id = (n, a_1, \id_{P_{A_2}})$, and likewise we define $\id \timesk (n,
  % a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.
\end{definition}

It is then easy to verify that the Kleisli arrow action on perturbation is well-defined 
in that for any $m \in M_{UB}$, we have
%
\[ \ptb_{U(A \arr B)}(\id \tok m) = \id \tok i_{UB}(m), \]
%
where the LHS is the Kleisli action on perturbation and the RHS is the Kleisli
action on morphisms. 
%
The proof uses the fact that the morphism $\delta$ commutes with all error
domain morphisms, which is a consequence of the definition of error domain
morphism.
%
The other verification involving the Kleisli arrow action is similar, as are the
two Kleisli product actions.

% %
% \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \i_B(b)) = 
%     \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
% %
% This can be checked by unfolding the definition of $\tok$; the proof uses the
% fact that $\delta_{UB}^*$ commutes with computation morphisms.
% %

% Given a perturbation $q = (n, a) \in P_{FA}$ we define $(n, a) \tok \id = (n, a,
% \id_{P_B}) \in P_{U(A \to B)}$. We need to show that
% %
% \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
% %
% By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes
% with computation morphisms.

@@

% %%%%%%%%%%%%%%%%%%%%%%
% % Model Construction %
% %%%%%%%%%%%%%%%%%%%%%%
% We now proceed with the construction of the model:

%     Define a step-2 model $\mathcal M'$ as follows:
%     \begin{itemize}
%       \item Value objects are pairs consisting of:
%       \begin{itemize}
%         \item A value object $A$ in $\vf$.
%         \item A monoid $P_A$ of perturbations
%         and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.

%       \end{itemize}

%       \item Computation objects are pairs consisting of:
%       \begin{itemize}
%         \item A computation object $B$ in $\ef$.
%         \item A monoid $P_B$ of perturbations and a monoid homomorphism
%         $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
%       \end{itemize}

%       \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively.
%       %, i.e.,
%       % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
%       %
%       % and likewise for computations.
